Title Text

### 2007-03-22 matrix/tagless-final.tex
a bit more write-up

this time with HOAS rather than deBruin indices. The language is close 
to the language of Xi03, but without the polymorphic lift and more 
constants so we can write better examples, enough to express
Fibonacci, factorial, and power.

How to make encoding flexible: abstract the interpreter
Haskell: type classes, OCaml: functors (give the signature of
Symantics without the 'sv part)

We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
the term `constructor' functions are defined as methods in a type class
Symantics (the name means that the class interface gives the syntax for
the source language and its instances give the semantics).
	class Symantics r where
	  int :: Int -> r Int
	  add :: r Int -> r Int -> r Int
	  lam :: (r a -> r b) -> r (a->b)
	  app :: r (a->b) -> r a -> r b
	  fix :: (r a -> r a) -> r a
	  ...
So, Haskell infers the type "Symantics r => r Int" for the example term
encoding above.

@@

@@

>>>>

this time with HOAS rather than deBruijn indices. The language is close 
to the language of Xi03, but without the polymorphic lift and more 
constants so we can write better examples, enough to express
Fibonacci, factorial, and power.

How to make encoding flexible: abstract the interpreter
We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
the term `constructor' functions are defined as methods in a type class
Symantics (the name means that the class interface gives the syntax for
the source language and its instances give the semantics).

\begin{code}
class Symantics repr where
    int :: Int -> repr Int                -- int literal
    bool :: Bool -> repr Bool             -- bool literal

    lam :: (repr a -> repr b) -> repr (a->b)
    app :: repr (a->b) -> repr a -> repr b
    fix :: (repr a -> repr a) -> repr a

    add :: repr Int -> repr Int -> repr Int
    mul :: repr Int -> repr Int -> repr Int
    if_ :: repr Bool -> repr a -> repr a -> repr a
    leq :: repr Int -> repr Int -> repr Bool
\end{code}

The previous test now reads |app (lam (\x -> x)) (bool True)| and the
inferred type is |Symantics repr => repr Bool|.

@@

\begin{code}
testpowfix () = lam (\x ->
                      fix (\self -> lam (\n ->
                        if_ (leq n (int 0)) (int 1)
                            (mul x (app self (add n (int (-1))))))))
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code}

OCaml: functors (give the signature of
Symantics without the 'sv part)

\begin{code}
module type Symantics = sig
  type ('c,'dv) repr
  val int  : int  -> ('c,int) repr
  val bool : bool -> ('c,bool) repr
  val add  : ('c,int) repr -> ('c,int) repr -> ('c,int) repr
  ...
  val if_ : ('c,bool) repr ->
             (unit -> ('c,'da) repr) ->
             (unit -> ('c,'da) repr) -> ('c,'da) repr

  val lam : (('c,'da) repr -> ('c,'db) repr) 
          -> ('c,'da->'db) repr
  val app : ('c,'da->'db) repr
    -> ('c,'da) repr -> ('c,'db) repr
  val fix : (('c,'da->'db) repr -> ('c,'da->'db) repr) 
            -> ('c,'da->'db) repr
end
\end{code}

Some differences are in the additional type parameter |c| (the
environment classifier, Walid03) to be used later when we get to code
generation (required by MetaOCaml, part of the type of the
code). MetaOCaml is a call-by-value language, and thus we have to eta-expand
the fix-point combinator and encapsulate the branches of the |if_|
into thunks to prevent their evaluation.

\begin{code}
module EX(S: Symantics) = struct
 open S
 let testpowfix () = lam (fun x ->
                      fix (fun self -> lam (fun n ->
                        if_ (leq n (int 0)) (fun () -> int 1)
                            (fun () -> mul x (app self (add n (int (-1))))))))

 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
The unit in |testpowfix| is for the sake of value restrcition (which
interferes with the environment classifier inference).

Thus the DSL expression has the type of the functor, from |Symantics|
to the value.

@@

Self-interpretation in our framework is
a bit tricky due to a need for polymorphism.  An encoded term should
be polymorphic in the interpreter type-constructor "r".  Further, the
functions "lam", "app", etc. above should be polymorphic in the object
types "a" and "b".  Thus the object language seems to require rank-2
polymorphism; and so type annotations will be inevitable.

To neatly bypass the need for rank-2 polymorphism, we use the notion
of let-polymorphism and a `hole', as follows.  A partial evaluator (or
a compiler or interpreter) is an evaluation context that (let-)binds
variables named "app", "lam", etc.  For the object language to
take advantage of let-bound polymorphism in the metalanguage, it
is crucial that we encode object "let" to meta "let".  We achieve
self-interpretation in that the following evaluation context in the
object language defines an interpreter for encoded object terms.
\begin{code}
    let lam = \f -> f in
    let app = \f x -> f x in
    let add = \m n -> m + n in
    let int = \i -> i in
    [hole]
\end{code}
The obvious notion of optimality in this setup is easy to achieve.  For
example, given the context above, our partial evaluator turns
\begin{code}
    app (lam (\x -> add x (int 1))) (int 3)
\end{code}
into the number 4, as desired.  We can also encode the self-interpreter
above as the following evaluation context in the object language.
\begin{code}
    let lam = lam (\f -> f) in
    let app = lam (\f -> lam (\x -> app f x)) in
    let add = lam (\m -> lam (\n -> add m n)) in
    let int = lam (\i -> i) in
    [hole]
\end{code}
The "let" and the hole "[hole]" are meta-constructions, which must map to
themselves in any interpretation.
